Space-time structure may be topological and not geometrical

### 2020-05-24 Proceedings/2019-ICNFP/Carcassi2019ICNFPProceedings.tex
Fixing date
Addressing reviewers comments

\address{Physics Department, University of Michigan, 450 Church Street\\
	Ann Arbor, MI 48109-1040,
	United States}
\ead{carcassi@umich.edu}
\vspace{10pt}
\begin{indented}
\item[]February 2019
\end{indented}

\begin{abstract}
In a previous effort we have created a framework that explains why topological structures naturally arise within a scientific theory; namely, they capture the requirements of experimental verification. This is particularly interesting because topological structures are at the foundation of geometrical structures, which play a fundamental role within modern mathematical physics. In this talk we will show a set of necessary and sufficient conditions under which those topological structures lead to real quantities and manifolds, which are a typical requirement for geometry. These conditions will provide a physically meaningful procedure that is the physical counter-part of the use of Dedekind cuts in mathematics. We then show that those conditions are unlikely to be met at Planck scale, leading to a breakdown of the concept of ordering. This would indicate that the mathematical structures required to describe space-time at that scale, while still topological, may not be geometrical.
\end{abstract}

@@

For those interested in the ultimate structure of space-time these insights naturally lead to the following line of inquiry. Before we want to understand whether space-time is geometrically truly a Riemmanian manifold, we need to understand whether space-time is topologically truly a manifold. A manifold, simply put, is a set of possible cases $X$ that can be experimentally identified by a set of real values. So, the question is: when can a set of possible cases, or possibilities for short, be experimentally identified by a real value? To be clear, the question is not when the results of our measurements are real numbers. The answer to this is simple: never. Our measurements are always finite precision. The question is when the cases we want to be able to distinguish are parameterized by the real line.

@@

If we think at how quantities are measured, the general idea is that we are able to define references with preset amounts, and then compare our object to several references to find the ones that bound the value from above and below. For example, a ruler is a series of marks and measuring the position with it means finding the closest one. A clock is a series of ticks and measuring time means noting the tick right before the event. A balance scale compares the weight of an object to a few known ones. Intuitively, we can understand that a quantity is continuous if, in line of principle, we can prepare references ever closer to each other. Our aim is to make this notion precise, so we have a practical/operational definition of what we mean experimentally by a continuous quantity.

@@

@@

Using that strategy, we reached two major findings. The first major finding is that \emph{most of the conditions are not required by the real numbers specifically, but they are required by ordered quantities in general}. That is, the set of reference has to satisfy a few conditions just to be sure that the cases they distinguish experimentally are ordered (i.e.~can be thought to be all being one after the other). The conditions are the following.
\begin{itemize}
	\item The references must be strict, meaning that the cases before, on and after are mutually exclusive. In practice, if objects have an extent, this must be much smaller than the extent of the reference, such that it will always be considered wholly before, on or after the reference.
	\item The references must be aligned, meaning their before and after statements identify incremental regions. In practice, this means we can always distinguish between them and they remain with constant before/after relationships (e.g.~they do not fluctuate). 
	\item The references must be refinable, meaning we can always resolve overlaps and fill in the whole space. For example, if an object can be between two references, then we must be able to put a new reference between those two references as well. If we have two references that overlap, then we must be able to find other references that fit within that overlap, to tell what is before and after.
\end{itemize}
These are the conditions one must require to have any linear order, either discrete (which can be labeled by integers), continuous (which can be labeled by real numbers) or any other. The divergence between real and integer quantities lies in how many references one can put in between two references. In the integer case, we can only fit finitely many references. In the real case, we can fit infinitely many.

@@

The inability to create an ordering at the finest scale means we are not able to define other geometrical quantities, such as distances and angles. In this light, the ultimate structure of space-time may not be geometrical but it will still be topological, as we will still need to describe what is experimentally accessible.

@@

@@

What does this requirement entail in practice? It means that we can place a reference wherever we want and can shrink its extent such that it occupies only one possibility.

@@

>>>>

\address{Physics Department, University of Michigan, 450 Church Street\\
	Ann Arbor, MI 48109-1040,
	United States}
\ead{carcassi@umich.edu}
\vspace{10pt}
\begin{indented}
\item[]February 2020
\end{indented}

\begin{abstract}
In a previous effort we have created a framework that explains why topological structures naturally arise within a scientific theory; namely, they capture the requirements of experimental verification. This is particularly interesting because topological structures are at the foundation of geometrical structures, which play a fundamental role within modern mathematical physics. In this paper we will show a set of necessary and sufficient conditions under which those topological structures lead to real quantities and manifolds, which are a typical requirement for geometry. These conditions will provide a physically meaningful procedure that is the physical counter-part of the use of Dedekind cuts in mathematics. We then show that those conditions are unlikely to be met at Planck scale, leading to a breakdown of the concept of ordering. This would indicate that the mathematical structures required to describe space-time at that scale, while still topological, may not be geometrical.
\end{abstract}

@@

For those interested in the ultimate structure of space-time these insights naturally lead to the following line of inquiry. Before we want to understand whether space-time is geometrically truly a Riemannian manifold, we need to understand whether space-time is topologically truly a manifold. A manifold, simply put, is a set of possible cases $X$ that can be experimentally identified by a set of real values. So, the question is: when can a set of possible cases, or possibilities for short, be experimentally identified by a real value? To be clear, the question is not when the results of our measurements are real numbers. The answer to this is simple: never. Our measurements are always finite precision. The question is when the cases we want to be able to distinguish are parameterized by the real line.

We will concentrate on a single real line (i.e.~total ordering) and not address higher dimensional spaces specifically. Since space-time in relativity is four dimensional, it would seem we are ignoring the most interesting cases, but it is not so. The issue is that time itself plays a special role: in every local frame, we must be able to use the time coordinate as the affine parameter for the evolution of a particle by writing $x^i=x^i(x^0)$. As we mentioned before, functions need to be, at the very least, topologically continuous or they would break the notion of experimental verifiability (i.e.~verifiable statements would not be mapped to other verifiable statements). This means that in any theory that includes time evolution, the topology one gives to time will severely constrain the topology of the space within which time evolution takes place. Moreover, in any theory of space-time we have the additional constraint that time and spatial coordinates can be mixed. Therefore the argument will work in reverse as well: the topology one gives to space has to be the same as the one of time because one may use spatial distance as a time parameter. For example, two spaceships drifting at constant velocity may use their spatial separation as a clock.

@@

If we think about how quantities are measured, the general idea is that we are able to define references with preset amounts, and then compare our object to several references to find the ones that bound the value from above and below. For example, a ruler is a series of marks and measuring the position with it means finding the closest one. A clock is a series of ticks and measuring time means noting the tick right before the event. A balance scale compares the weight of an object to a few known ones. Intuitively, we can understand that a quantity is continuous if, in line of principle, we can prepare references ever closer to each other. Our aim is to make this notion precise, so we have a practical/operational definition of what we mean experimentally by a continuous quantity.

@@

The idea is that the quantities themselves are not a priori objects, but a construction built upon a set of references that form our system of measurement. A reference frame in space-time, then, is also a construction consisting of fixed elements (e.g. the stars, the borders of the experiment table, the elements of a timing system in a particle accelerator, ...) and signals exchanged among them. The topology of space-time is then an idealized characterization of the set of all possible such constructions, that is, all possible reference systems. Our goal is to understand the extent and the limitation of such idealization.

@@

Using that strategy, we reached two major findings. The first major finding is that \emph{most of the conditions are not required by the real numbers specifically, but they are required by ordered quantities in general}. That is, the set of references has to satisfy a few conditions just to be sure that the cases they distinguish experimentally are ordered (i.e.~can be thought to be all being one after the other). The conditions are the following.
\begin{itemize}
	\item The references must be strict, meaning that the cases before, on and after are mutually exclusive. In practice, if objects have an extent, this must be much smaller than the extent of the reference, such that it will always be considered wholly before, on or after the reference.
	\item The references must be aligned, meaning their before and after statements identify incremental regions. In practice, this means we can always distinguish between them and they remain with constant before/after relationships (e.g.~they do not fluctuate). 
	\item The references must be refinable, meaning we can always resolve overlaps and fill in the whole space. For example, if an object can be between two references, then we must be able to put a new reference between those two references as well. If we have two references that overlap, then we must be able to find other references that fit within that overlap, to tell what is before and after.
\end{itemize}
These are the conditions one must require to have any linear order, either discrete (which can be labeled by integers), continuous (which can be labeled by real numbers) or any other. The divergence between real and integer quantities lies in how many references one can put in between two references. In the integer case, we can only fit finitely many references. In the real case, we can fit infinitely many.

@@

The inability to create an ordering at the finest scale means we are not able to define other geometrical quantities, such as distances and angles. In this light, the ultimate structure of space-time may not be geometrical but it will still be topological, as we will still need to describe what is experimentally accessible. At large scale, geometrical structure will need to emerge to recover the established theories. The point is that these large scale geometric structures cannot emerge from other geometrical structures, as these would necessarily suffer from the same fundamental problems.

@@

To give a better understanding of how these definitions work in a multidimensional setting, let us see how they apply to a Cartesian frame with coordinates $(x, y, z)$. A reference for the $x$ coordinate would be something like $r_1=$(``$x < 0$'',``$0 \leq x \leq 2$'', ``$ x > 2$''). Note that it partitions the whole space in three disjoint regions. The on case acts as a divider and extends throughout $y$ and $z$. The reference $r_2=$(``$y < 0$'',``$0 \leq y \leq 1$'', ``$y > 1$'') is not aligned with $r_1$ as something can be before $r_1$ and still be before, on or after $r_2$. On the other hand, $r_3=$(``$x < -1$'',``$-1 \leq x \leq 1$'', ``$x > 1$'') is aligned with $r_1$ even though it is not either before or after $r_1$ because the on regions overlap. Note that these references would not have such a simple expression in, say, spherical coordinates. Yet the regions themselves are coordinate independent and, therefore, the relationships between them are too.

What happens is that each reference frame would have its own set of references that can be nicely expressed using its coordinates. The before/after relationships expressed using one set of references may not be simply expressible using another, precisely because they are not aligned with each other. Relativistically, the references used for time by one frame will not be aligned with the ones of a boosted frame: the coordinate surfaces do not partition (i.e. do not foliate) space-time in the same way. Reference alignment essentially captures these various requirements with one simple formal definition.

@@

What does this requirement entail in practice? It means that we can place a reference wherever we want and can shrink its extent such that it occupies only one possibility. If we take the traditional manifold structure of space-time literally, this would mean having at our disposal ever shrinking references that can be placed anywhere in space and time. For example, it would require being able to create a timing system with as many well separated pulses as desired distributed to as many places as desired. All of this, without changing the nature of the process we are studying.

@@

%Note that references have to be physically implemented somehow, either by placing massive objects or sending signals at the reference points, which will affect the object under study, at the very least by gravitational effects. But for many applications such reference-induced effects can be neglected. and the process can be repeated sufficiently well that, in practice, we can assume

%Riemannian
%set of reference (missing s)
%think at
